### 2016-09-15 revision/preliminaries.tex
typos of reviewer 2 and his simplified proof

\textsc{Mirror Descent} (MD) is a generic algorithm similar to \textsc{FTRL}
but quite different in details. The algorithm is stated as
Algorithm~\ref{algorithm:mirror-descent-varying-regularizer}. The algorithm is
parametrized by a sequence $\{R_t\}_{t=0}^\infty$ of convex functions $R_t:K
\to \R$ called \emph{regularizers}. Each regularizer $R_t$ can depend on past
loss vectors $\ell_1, \ell_2, \dots, \ell_t$ in an arbitrary way. If $R_t$ is
not differentiable,\footnote{Note that this can happen even when $R_t$ is a
restriction of a differentiable function defined on a superset of $K$.  If $K$
is bounded and closed, $R_t$ fails to be differentiable at the boundary of $K$.
If $K$ is a subset of an affine subspace of a dimension smaller than the
dimension of $V$, then $R_t$ fails to be differentiable everywhere.} the
Bregman divergence, $\Breg_{R_t}(u,v) = R_t(u) - R_t(v) - \langle \grad R_t(v),
u - v \rangle$ needs to be defined. This is done by choosing a subgradient map
$\grad R_t:K \to V$, i.e., a function such that $\grad R_t(w)$ is a subgradient
of $R_t$ at a point $w$. If $R_t$ is a restriction of a differentiable function
$R'_t$, it is convenient to define $\grad R_t(w) = \grad R'_t(w)$ for all $w
\in K$. The following lemma bounds regret of \textsc{MD}.  Proof can be found
in~\cite{Rakhlin-Sridharan-2009,Duchi-Shalev-Shwartz-Singer-Tewari-2010}.  For
completeness we give a proof in~\ref{section:mirror-descent-proofs}.

>>>>

\textsc{Mirror Descent} (MD) is a generic algorithm similar to \textsc{FTRL}
but quite different in the details. The algorithm is stated as
Algorithm~\ref{algorithm:mirror-descent-varying-regularizer}. The algorithm is
parametrized by a sequence $\{R_t\}_{t=0}^\infty$ of convex functions $R_t:K
\to \R$ called \emph{regularizers}. Each regularizer $R_t$ can depend on past
loss vectors $\ell_1, \ell_2, \dots, \ell_t$ in an arbitrary way. If $R_t$ is
not differentiable,\footnote{Note that this can happen even when $R_t$ is a
restriction of a differentiable function defined on a superset of $K$.  If $K$
is bounded and closed, $R_t$ fails to be differentiable at the boundary of $K$.
If $K$ is a subset of an affine subspace of a dimension smaller than the
dimension of $V$, then $R_t$ fails to be differentiable everywhere.} the
Bregman divergence, $\Breg_{R_t}(u,v) = R_t(u) - R_t(v) - \langle \grad R_t(v),
u - v \rangle$ needs to be defined. This is done by choosing a subgradient map
$\grad R_t:K \to V$, i.e., a function such that $\grad R_t(w)$ is a subgradient
of $R_t$ at any point $w$. If $R_t$ is a restriction of a differentiable function
$R'_t$, it is convenient to define $\grad R_t(w) = \grad R'_t(w)$ for all $w
\in K$. The following lemma bounds the regret of \textsc{MD}.  A proof can be found
in~\cite{Rakhlin-Sridharan-2009,Duchi-Shalev-Shwartz-Singer-Tewari-2010}.  For
completeness we give a proof in~\ref{section:mirror-descent-proofs}.

### 2016-09-15 revision/solo-ftrl.tex
typos of reviewer 2 and his simplified proof

In the following, we propose a much simpler and better approach.  We propose to
use Algorithm~\ref{algorithm:ftrl-varying-regularizer} with the regularizer
\begin{equation}
\label{equation:solo-ftrl-regularizer}
R_t(w) = R(w) \sqrt{\sum_{i=1}^{t-1} \|\ell_i\|_*^2} \; ,
\end{equation}
where $R:K \to \R$ is any strongly convex function. Through a refined analysis,
we show that the regularizer suffices to obtain an optimal regret bound for any
decision set, bounded or unbounded.  We call such variant \textsc{Scale-free
Online Linear Optimization FTRL} algorithm (\textsc{SOLO FTRL}).  Our main
result is Theorem~\ref{theorem:regret-solo-ftrl} below, which is proven in
Section~\ref{section:solo-ftrl-regret-bound}.

The regularizer~\eqref{equation:solo-ftrl-regularizer} does not uniquely define
the \textsc{FTRL} minimizer $w_t = \argmin_{w \in K} R_t(w)$ when
$\sqrt{\sum_{i=1}^{t-1} \|\ell_i\|_*^2}$ is zero. This happens if $\ell_1,
\ell_2, \dots, \ell_{t-1}$ are all zero (and in particular for $t=1$).  In that
case, we define the $w_t = \argmin_{w \in K} R(w)$ which is consistent with
$w_t = \lim_{a \to 0^+}  \argmin_{w \in K} a R(w)$.

@@

\begin{proof}[Proof of Theorem~\ref{theorem:regret-solo-ftrl}]
Let $\eta_t = \frac{1}{\sqrt{\sum_{i=1}^{t-1} \|\ell_i\|_*^2}}$, hence $R_t(w)
= \frac{1}{\eta_t} R(w)$.  We assume without loss of generality that
$\|\ell_t\|_* > 0$ for all $t$, since otherwise we can remove all rounds $t$
where $\ell_t = 0$ without affecting regret and the predictions of the
algorithm on the remaining rounds.  By Lemma~\ref{lemma:generic-regret-bound},
\begin{align*}
\Regret_T(u)
& \le \frac{1}{\eta_{T+1}} R(u) + \sum_{t=1}^T \left( \Breg_{R_t^*}(-L_t, -L_{t-1}) - R_t^*(-L_t) + R_{t+1}^*(-L_t) \right) \; .
\end{align*}
We upper bound the terms of the sum in two different ways.
First, by Proposition~\ref{proposition:conjugate-properties}, we have
$$
\Breg_{R_t^*}(-L_t, -L_{t-1}) - R_t^*(-L_t) + R_{t+1}^*(-L_t)
\le \Breg_{R_t^*}(-L_t, -L_{t-1})
\le \frac{\eta_t \|\ell_t\|_*^2}{2\lambda} \; .
$$
Second, we have
\begin{align*}
\Breg_{R_t^*}&(-L_t, -L_{t-1}) - R_t^*(-L_t) + R_{t+1}^*(-L_t) \\
& = \Breg_{R_{t+1}^*}(-L_t, -L_{t-1}) + R^*_{t+1}(-L_{t-1}) - R_t^*(-L_{t-1}) \\
& \qquad + \langle \nabla R_t^*(-L_{t-1})-\nabla R_{t+1}^*(-L_{t-1}), \ell_t \rangle  \\
& \le \frac{\eta_{t+1} \|\ell_t\|_*^2}{2\lambda} + \| \nabla R_t^*(-L_{t-1})-\nabla R_{t+1}^*(-L_{t-1})\| \cdot \|\ell_t\|_* \\
& = \frac{\eta_{t+1} \|\ell_t\|_*^2}{2\lambda} + \| \nabla R^*(- \eta_{t} L_{t-1})-\nabla R^*(- \eta_{t+1} L_{t-1})\| \cdot \|\ell_t\|_* \\
& \le \frac{\eta_{t+1} \|\ell_t\|_*^2}{2\lambda} + \min\left\{\frac{1}{\lambda} \|L_{t-1}\|_* \left(\eta_{t} - \eta_{t+1} \right), D\right\} \|\ell_t\|_* \; ,
\end{align*}
where in the first inequality we have used the fact that $R^*_{t+1}(-L_{t-1})
\le R_t^*(-L_{t-1})$, H\"older's inequality, and
Proposition~\ref{proposition:conjugate-properties}.  In the second inequality
we have used properties 5 and 7 of
Proposition~\ref{proposition:conjugate-properties}. Using the definition of
$\eta_{t+1}$ we have
\begin{align*}
\frac{\|L_{t-1}\|_* (\eta_{t} -\eta_{t+1})}{\lambda}
\le \frac{ \|L_{t-1}\|_*}{\lambda \sqrt{\sum_{i=1}^{t-1} \|\ell_i\|_*^2}}
\le \frac{ \sum_{i=1}^{t-1} \|\ell_i\|_*}{\lambda \sqrt{\sum_{i=1}^{t-1} \|\ell_i\|_*^2}}
\le \frac{\sqrt{t-1}}{\lambda}
\le \frac{\sqrt{T-1}}{\lambda}.
\end{align*}
Denoting by $H=\min\left\{\frac{\sqrt{T-1}}{\lambda},D\right\}$ we have
\begin{align*}
& \Regret_T(u)
\le \frac{1}{\eta_{T+1}} R(u) + \sum_{t=1}^T \min\left\{ \frac{\eta_t \|\ell_t\|_*^2}{2\lambda}, \ H \|\ell_t\|_* + \frac{\eta_{t+1} \|\ell_t\|_*^2}{2\lambda}  \right\} \\
& \le \frac{1}{\eta_{T+1}} R(u) + \frac{1}{2\lambda} \sum_{t=1}^T  \eta_{t+1} \|\ell_t\|_*^2 + \frac{1}{2\lambda} \sum_{t=1}^T \min\left\{ \eta_t \|\ell_t\|_*^2, \ 2 \lambda H \|\ell_t\|_* \right\} \\
& = \frac{1}{\eta_{T+1}} R(u) + \frac{1}{2\lambda} \sum_{t=1}^T  \frac{\|\ell_t\|_*^2}{\sqrt{\sum_{i=1}^t \|\ell_i\|_*^2}} + \frac{1}{2 \lambda} \sum_{t=1}^T \min\left\{ \frac{\|\ell_t\|_*^2}{\sqrt{\sum_{i=1}^{t-1} \|\ell_i\|_*^2}}, \ 2 \lambda H \|\ell_t\|_* \right\} \; .
\end{align*}
We bound each of the three terms separately. By definition of $\eta_{T+1}$, the
first term is $\frac{1}{\eta_{T+1}} R(u) = R(u) \sqrt{\sum_{t=1}^T
\|\ell_t\|_*^2}$.  We upper bound the second term using
Lemma~\ref{lemma:sum-of-square-roots-inverses} as
$$
\frac{1}{2\lambda} \sum_{t=1}^T  \frac{\|\ell_t\|_*^2}{\sqrt{\sum_{i=1}^t \|\ell_i\|_*^2}}
\le \frac{1}{\lambda} \sqrt{\sum_{t=1}^T \|\ell_t\|_*^2} \; .
$$
Finally, by Lemma~\ref{lemma:useful} we upper bound the third term as
$$
\frac{1}{2 \lambda} \sum_{t=1}^T \min\left\{ \frac{\|\ell_t\|_*^2}{\sqrt{\sum_{i=1}^{t-1} \|\ell_i\|_*^2}}, \ 2 \lambda \|\ell_t\|_* H \right\}
\le 3.5 H \max_{t \le T} \|\ell_t\|_* + \frac{1.75}{\lambda} \sqrt{\sum_{t=1}^T \|\ell_t\|_*^2} \; .
$$
Putting everything together gives the stated bound.
\end{proof}

>>>>

In the following, we propose a much simpler and better approach.  We propose to
use Algorithm~\ref{algorithm:ftrl-varying-regularizer} with the regularizer
\begin{equation}
\label{equation:solo-ftrl-regularizer}
R_t(w) = R(w) \sqrt{\sum_{i=1}^{t-1} \|\ell_i\|_*^2} \; ,
\end{equation}
where $R:K \to \R$ is any strongly convex function. Through a refined analysis,
we show that this regularizer suffices to obtain an optimal regret bound for any
decision set, bounded or unbounded.  We call this variant \textsc{Scale-free
Online Linear Optimization FTRL} algorithm (\textsc{SOLO FTRL}).  Our main
result is Theorem~\ref{theorem:regret-solo-ftrl} below, which is proven in
Section~\ref{section:solo-ftrl-regret-bound}.

The regularizer~\eqref{equation:solo-ftrl-regularizer} does not uniquely define
the \textsc{FTRL} minimizer $w_t = \argmin_{w \in K} R_t(w)$ when
$\sqrt{\sum_{i=1}^{t-1} \|\ell_i\|_*^2}$ is zero. This happens if $\ell_1,
\ell_2, \dots, \ell_{t-1}$ are all zero (and in particular for $t=1$).  In that
case, we define $w_t = \argmin_{w \in K} R(w)$ which is consistent with
$w_t = \lim_{a \to 0^+}  \argmin_{w \in K} a R(w)$.

@@

\begin{proof}[Proof of Theorem~\ref{theorem:regret-solo-ftrl}]
Let $\eta_t = \frac{1}{\sqrt{\sum_{i=1}^{t-1} \|\ell_i\|_*^2}}$, hence $R_t(w)
= \frac{1}{\eta_t} R(w)$.  We assume without loss of generality that
$\|\ell_t\|_* > 0$ for all $t$, since otherwise we can remove all rounds $t$
where $\ell_t = 0$ without affecting the regret and the predictions of the
algorithm on the remaining rounds.  By Lemma~\ref{lemma:generic-regret-bound},
\begin{align*}
\Regret_T(u)
& \le \frac{1}{\eta_{T+1}} R(u) + \sum_{t=1}^T \left( \Breg_{R_t^*}(-L_t, -L_{t-1}) - R_t^*(-L_t) + R_{t+1}^*(-L_t) \right) \; .
\end{align*}
We upper bound the terms of the sum in two different ways.
First, by Proposition~\ref{proposition:conjugate-properties}, we have
$$
\Breg_{R_t^*}(-L_t, -L_{t-1}) - R_t^*(-L_t) + R_{t+1}^*(-L_t)
\le \Breg_{R_t^*}(-L_t, -L_{t-1})
\le \frac{\eta_t \|\ell_t\|_*^2}{2\lambda} \; .
$$
Second, we have
\begin{align*}
\Breg_{R_t^*}&(-L_t, -L_{t-1}) - R_t^*(-L_t) + R_{t+1}^*(-L_t) \\
& = \Breg_{R_{t+1}^*}(-L_t, -L_{t-1}) + R^*_{t+1}(-L_{t-1}) - R_t^*(-L_{t-1}) \\
& \qquad + \langle \nabla R_t^*(-L_{t-1})-\nabla R_{t+1}^*(-L_{t-1}), \ell_t \rangle  \\
& \le \frac{\eta_{t+1} \|\ell_t\|_*^2}{2\lambda} + \| \nabla R_t^*(-L_{t-1})-\nabla R_{t+1}^*(-L_{t-1})\| \cdot \|\ell_t\|_* \\
& = \frac{\eta_{t+1} \|\ell_t\|_*^2}{2\lambda} + \| \nabla R^*(- \eta_{t} L_{t-1})-\nabla R^*(- \eta_{t+1} L_{t-1})\| \cdot \|\ell_t\|_* \\
& \le \frac{\eta_{t+1} \|\ell_t\|_*^2}{2\lambda} + \min\left\{\frac{1}{\lambda} \|L_{t-1}\|_* \left(\eta_{t} - \eta_{t+1} \right), D\right\} \|\ell_t\|_* \; ,
\end{align*}
where in the first inequality we have used the fact that $R^*_{t+1}(-L_{t-1})
\le R_t^*(-L_{t-1})$, H\"older's inequality, and
Proposition~\ref{proposition:conjugate-properties}.  In the second inequality
we have used properties 5 and 7 of
Proposition~\ref{proposition:conjugate-properties}. Using the definition of
$\eta_{t+1}$ we have
\begin{align*}
\frac{\|L_{t-1}\|_* (\eta_{t} -\eta_{t+1})}{\lambda}
\le \frac{ \|L_{t-1}\|_*}{\lambda \sqrt{\sum_{i=1}^{t-1} \|\ell_i\|_*^2}}
\le \frac{ \sum_{i=1}^{t-1} \|\ell_i\|_*}{\lambda \sqrt{\sum_{i=1}^{t-1} \|\ell_i\|_*^2}}
\le \frac{\sqrt{t-1}}{\lambda}
\le \frac{\sqrt{T-1}}{\lambda}.
\end{align*}
Denoting by $H=\min\left\{\frac{\sqrt{T-1}}{\lambda},D\right\}$ we have
\begin{align*}
& \Regret_T(u)
\le \frac{1}{\eta_{T+1}} R(u) + \sum_{t=1}^T \min\left\{ \frac{\eta_t \|\ell_t\|_*^2}{2\lambda}, \ H \|\ell_t\|_* + \frac{\eta_{t+1} \|\ell_t\|_*^2}{2\lambda}  \right\} \\
& \le \frac{1}{\eta_{T+1}} R(u) + \frac{1}{2\lambda} \sum_{t=1}^T  \eta_{t+1} \|\ell_t\|_*^2 + \frac{1}{2\lambda} \sum_{t=1}^T \min\left\{ \eta_t \|\ell_t\|_*^2, \ 2 \lambda H \|\ell_t\|_* \right\} \\
& = \frac{1}{\eta_{T+1}} R(u) + \frac{1}{2\lambda} \sum_{t=1}^T  \frac{\|\ell_t\|_*^2}{\sqrt{\sum_{i=1}^t \|\ell_i\|_*^2}} + \frac{1}{2 \lambda} \sum_{t=1}^T \min\left\{ \frac{\|\ell_t\|_*^2}{\sqrt{\sum_{i=1}^{t-1} \|\ell_i\|_*^2}}, \ 2 \lambda H \|\ell_t\|_* \right\} \; .
\end{align*}
We bound each of the three terms separately. By definition of $\eta_{T+1}$, the
first term is $\frac{1}{\eta_{T+1}} R(u) = R(u) \sqrt{\sum_{t=1}^T
\|\ell_t\|_*^2}$.  We upper bound the second term using
Lemma~\ref{lemma:sum-of-square-roots-inverses} as
$$
\frac{1}{2\lambda} \sum_{t=1}^T  \frac{\|\ell_t\|_*^2}{\sqrt{\sum_{i=1}^t \|\ell_i\|_*^2}}
\le \frac{1}{\lambda} \sqrt{\sum_{t=1}^T \|\ell_t\|_*^2} \; .
$$
Finally, by Lemma~\ref{lemma:useful} we upper bound the third term as
$$
\frac{1}{2 \lambda} \sum_{t=1}^T \min\left\{ \frac{\|\ell_t\|_*^2}{\sqrt{\sum_{i=1}^{t-1} \|\ell_i\|_*^2}}, \ 2 \lambda \|\ell_t\|_* H \right\}
\le 3.5 H \max_{t \le T} \|\ell_t\|_* + \frac{1.75}{\lambda} \sqrt{\sum_{t=1}^T \|\ell_t\|_*^2} \; .
$$
Putting everything together gives the stated bound.
\end{proof}
